Six steps to a better relationship with your future self.

### 2011-03-24 workflow.Rnw
More work

\title{How to improve your relationship with your future self.}

\author{Jake Bowers\thanks{I owe many thanks to Mark Fredrickson,
    Brian Gaines, Mark Fredrickson, Kieran Healy, Kevin Quinn and Cara
    Wong for direct help on this document and to Mika LaVaque-Manty
    and Ben Hansen for many useful discussions on this topic. The
    source code for this document may be freely downloaded and
    modified from \url{https://github.com/jwbowers/workflow}.}}

@@

\usepackage[noae,nogin]{Sweave} %Declaring this explicitly so that I can modify the Sweave environment here rather in Sweave.sty which will be c
\DefineVerbatimEnvironment{Sinput}{Verbatim} {xleftmargin=.5em,fontsize=\small}
\DefineVerbatimEnvironment{Soutput}{Verbatim}{xleftmargin=.5em,fontsize=\small}
\DefineVerbatimEnvironment{Scode}{Verbatim}  {xleftmargin=.5em,fontsize=\small}
\fvset{listparameters={\setlength{\topsep}{0pt}}}
\renewenvironment{Schunk}{\vspace{\topsep}}{\vspace{\topsep}}
\SweaveOpts{keep.source=TRUE}  %Show comments
\fvset{fontsize=\small} %slightly smaller verbatim font

@@

This idea inspires a seminar paper in the spring. A conference
paper arises from the seminar paper in collaboration with another
student in 2012. A dissertation chapter descends from the conference
paper in 2013. Other dissertation chapters take up 2014. A submission
to a journal with the original co-author and a new collaborator
happens in 2015. Revision and resubmission wait until 2017 while
harried editors, reviewers and authors strive to balance teaching,
service, and life. By now, the three lucky collaborators work as
professors in three different universities. In 2018 a child is born
and a paper is published. The United Nations takes an interest in the
paper in 2019 and hosts a conference to discuss implications of the
research. In 2020 a first year graduate student in a coffee shop has
an idea that challenges the results in the now famous paper. What
would happen if the authors had controlled for X? Or run one more MCMC chain?
Or chosen a different likelihood function?  Will the United Nations
(now eager to act based the paper) make a wrong move?

The first author convenes a three way video conference with the other
collaborators during his homeward commute after putting his flying car
in auto-drive mode.\footnote{One assumes that video chatting during
  manual driving of flying cars will have been outlawed in his state
  in 2017.} The group must go back to the analyses. Which ones?  The
ones from 2011?  Or 2012? Or 2018? Where are the files? The next day,
one member of the group who has kept some hard-drives around out of
nostalgia finds some of the files.\footnote{This is the same guy who
  still owns cassette tapes and compact discs.} Now the checks and
re-analyses should be easy. Right? The student, now professor, should
remember the reason for those bits of code (or at least should
remember which series of mouse clicks were used to produce the numbers
for that crucial table --- and that mousing should not be time
consuming to redo exactly as they were done in 2011, or was it
2015?). Right? And, of course, the way Microsoft
Word/Stata/SPSS/R/LISREL stores files and the way that current
machines in 2020 read and write them is the same --- since Windows and
Mac OS X have always existed and will always continue to exist more or
less as they currently exist? Right? And the group knows exactly which
bit of code produced which table and which figure, right? And they
wrote their code following Nagler's Maxims \citep{nagler1995coding}
and King's Replication Standard \citep{king1995replication} right?

Right? Wrong? If the collaborators find themselves answering ``wrong''
to many of the questions posed here then reproducing the analyses will
take a lot of time if such reproduction is possible at all. This essay
provides some suggestions for practices which will make such
reproduction occur much more easily and quickly in the event that
famous papers require special scrutiny. Specifically, this piece aims
to amplify some of what we already ought to know from
\citet{nagler1995coding} and \citet{king1995replication}, and to
update some of those ideas given current practices, platforms, and
propensities.

@@

\begin{Verbatim}[fontsize=\small]
thedata<-read.csv("Data/thedata-15-03-2011.csv") ## Read the data
pdf('fig1.pdf') ## begin writing to the pdf file
please-plot(outcome by explanatory using thedata. red lines please.)
dev.off() ## stop writing to the pdf file
\end{Verbatim}

Now, in the future if I wonder how ``that plot on page 10'' was
created: (1) ``that plot'' is from a file called \Verb+fig1.pdf+ and
\Verb+fig1.pdf+ was created in \Verb+fig1.R+. In a future where R still
exists, if I wanted to change the figure, I could edit the commands
quickly and easily to do so (since R commands are written in plain
text, and plain text is a format that will be around for a long time
to come).

\paragraph{Principle} Know the provenance of your results so that your
future self or current collaborators can quickly and easily reproduce
your work (and make changes).% --- to know what
% commands created which figures and tables.

\section{No data analyst is an island for long.}
Data analysis involves a long series of decisions. Each decision
requires justification.  The text itself will
justify the most important decisions. Some decisions will be too small
and technical for inclusion in the paper itself. These need to be
documented in the code itself \citep{nagler1995coding}.

\subsection{Comment your code.}
Comments --- unexecuted text inside of a script --- are a message to
your future self and other consumers of your work. Luckily, if you are
using a literate programming practice (ex. Sweave or odfWeave), you
can write paragraphs to surround your code as well as technical
comments in the code itself. If you are not using a strictly literate
programming practice then use whatever commenting protocol exists in
your analysis language. R, for example, considers text marked with
``\#'' as a comment. Here for example I comment about a linear model
run for a paper and also include comments about another model run for
due diligence.

<<reg1,echo=TRUE,results=verb>>=
##  Repeat the regression run for Table~\ref{tab:protest}
##  Do political rights predict protest reporting?
lm1 <- lm(protac2000 ~ I(Gini2004/100) + meanpr, data=good.df)
##  Q: Was it worthwhile controlling for political rights?
lm2 <- lm(protac2000~I(Gini2004/100),data=good.df)
coef(lm2)[2]-coef(lm1)[2]
##  A: Yes (around 3/4 an act of protest difference)
@

Messages left for your future self (or near-future others) help
retrace and justify the steps of your data-analytic journey in case
you ever need to revist them as the work moves from seminar paper to
conference paper to poster back to paper to dissertation and
onwards. Notice one another benefit of coding for an audience: we
learn by teaching. By assuming that others will look at your code, you
will be more likely to write clearer code, or perhaps even to learn
more about what you are doing as you do it.

>>>>

\title{Six steps to a better relationship with your future self.}

\author{Jake Bowers\thanks{I owe many thanks to Mark Fredrickson,
    Brian Gaines, Kieran Healy, Kevin Quinn and Cara
    Wong for direct help on this document and to Mika LaVaque-Manty
    and Ben Hansen for many useful discussions on this topic. The
    source code for this document may be freely downloaded and
    modified from \url{https://github.com/jwbowers/workflow}.}}

@@

\usepackage[noae,nogin]{Sweave} %Declaring this explicitly so that I can modify the Sweave environment here rather in Sweave.sty which will be c
\DefineVerbatimEnvironment{Sinput}{Verbatim} {xleftmargin=.5em,fontsize=\footnotesize}
\DefineVerbatimEnvironment{Soutput}{Verbatim}{xleftmargin=.5em,fontsize=\footnotesize}
\DefineVerbatimEnvironment{Scode}{Verbatim}  {xleftmargin=.5em,fontsize=\footnotesize}
\fvset{listparameters={\setlength{\topsep}{0pt}}}
\renewenvironment{Schunk}{\vspace{\topsep}}{\vspace{\topsep}}
\SweaveOpts{keep.source=TRUE}  %Show comments
\fvset{fontsize=\footnotesize} %slightly smaller verbatim font

@@

This idea inspires a seminar paper in the spring. A conference paper
arises from the seminar paper in collaboration with another student in
2012. A dissertation chapter descends from the conference paper in
2013. Other dissertation chapters take up 2014. A submission to a
journal with the original co-author and a new collaborator happens in
2015. Revision and resubmission wait until 2017 while harried editors,
reviewers and authors strive to balance research, teaching, service,
and life. By now, the three lucky collaborators work as professors in
three different universities. In 2018 a child is born and a paper is
published. The United Nations takes an interest in the paper in 2019
and hosts a conference to discuss implications of the research. In
2020 a first year graduate student in a coffee shop has an idea that
challenges the results in the now famous paper. What would happen if
the authors had controlled for X? Or included information now
available but missing in 2012?  Or chosen a different likelihood
function?  Will the United Nations (now eager to act based the paper)
make a wrong move?

The first author convenes a three way video conference with the other
collaborators during his homeward commute after putting his flying car
in auto-drive mode.\footnote{One assumes that video chatting during
  manual driving of flying cars will have been outlawed in his state
  by 2020.} The group must go back to the analyses. Which ones?  The
ones from 2011?  Or 2012? Or 2018? Where are the files? The next day,
one member of the group who has kept some hard-drives around out of
nostalgia finds some of the files.\footnote{This is the same guy who
  still owns cassette tapes and compact discs.} Now
re-analyses should be easy. Right? The student, now professor, should
remember the reason for those bits of code (or at least should
remember which series of mouse clicks were used to produce the numbers
for that crucial table --- and that mousing should not be time
consuming to redo exactly as it was done in 2011 \ldots or was it
2015?). Right? And, of course, the way Microsoft
Word/Stata/SPSS/R/LISREL understands files and the way that 
machines in 2020 read and write them is the same --- since Windows and
Mac OS X have always existed and will always continue to exist more or
less as they currently exist? Right? And the group knows exactly which
bit of code produced which table and which figure, right? And they
wrote their code following Nagler's Maxims \citep{nagler1995coding}
and King's Replication Standard \citep{king1995replication} right?

Right? Wrong? If the collaborators find themselves answering ``wrong''
to the questions posed here then reproducing, updating, or changing
the analyses will take a lot of time. If reproduction is hard to do,
then the reputations of the scholars will suffer and, more
importantly, the UN will be unable to implement a master plan for
world peace. This essay provides some suggestions for practices which
will make such reproduction occur much more easily and quickly in the
event that famous papers require special scrutiny. Specifically, this
piece aims to amplify some of what we already ought to know from
\citet{nagler1995coding} and \citet{king1995replication}, and to add
to some of those ideas given current practices, platforms, and
possibilities.

@@

\begin{Verbatim}[fontsize=\footnotesize]
thedata<-read.csv("Data/thedata-15-03-2011.csv") ## Read the data
pdf('fig1.pdf') ## begin writing to the pdf file
please-plot(outcome by explanatory using thedata. red lines please.)
please-add-a-line(using model1)
## Note to self: the following is no different from model1
## model2<-please-fit(outcome by explanatory+explanatory^2 using thedata
## summary(abs(fitted(model1)-fitted(model2)))
dev.off() ## stop writing to the pdf file
\end{Verbatim}

Now, in the future if I wonder how ``that plot on page 10'' was
created, I will know: (1) ``that plot'' is from a file called
\Verb+fig1.pdf+ and (2) \Verb+fig1.pdf+ was created in
\Verb+fig1.R+. In a future where R still exists, if I wanted to change
the figure, I could edit the commands quickly and easily to do so
(since R commands are written in
\href{http://en.wikipedia.org/wiki/Plain_text}{plain text}, and plain
text is a format that will be around as long as computer programmers
write computer programs). In a future where R does not exist, I will
at least be able to read the plain text R commands and use them to
write code in my new favorite statistical computing
language.\footnote{A most, since R is open source, I will be able to
  download R, download an old fashioned open-source operating systems
  (like Ubuntu 10), and run the old-fashioned statistical computing
  environment in the old-fashioned operating system in a virtual
  machine on my new-fashioned actual machine.}

\paragraph{Step 1} Know the provenance of your results so that your
future self or current collaborators can quickly and easily reproduce
your work (and make changes).% --- to know what
% commands created which figures and tables.

\section{No data analyst is an island for long.}
Data analysis involves a long series of decisions. Each decision
requires justification.  Some decisions will be too small and
technical for inclusion in the paper itself. These need to be
documented in the code itself \citep{nagler1995coding}. Paragraphs and
citations in the article itself will justify the most important
decisions.

\subsection{Code to communicate: Comment your code.}
Comments --- unexecuted text inside of a script --- are a message to
collaborators (including your future self) and other consumers of your
work. In the above code, I used comments to explain the lines to
readers unfamiliar with R and to remember that I had tried a different
specification but decided not to use it because adding the squared
term did not really change the predictions.\footnote{R considers text
  marked with ``\#'' as a comment.}

% Luckily, if you are
% using a literate programming practice (ex. Sweave or odfWeave), you
% can write paragraphs to surround your code as well as technical
% comments in the code itself. If you are not using a strictly literate
% programming practice then use whatever commenting protocol exists in
% your analysis language. R, for example, considers text marked with
% ``\#'' as a comment. Here for example I comment about a linear model
% run for a paper and also include comments about another model run for
% due diligence.

% <<reg1,echo=TRUE,results=verb>>=
% ##  Repeat the regression run for Table~\ref{tab:protest}
% ##  Do political rights predict protest reporting?
% lm1 <- lm(protac2000 ~ I(Gini2004/100) + meanpr, data=good.df)
% ##  Q: Was it worthwhile controlling for political rights?
% lm2 <- lm(protac2000~I(Gini2004/100),data=good.df)
% coef(lm2)[2]-coef(lm1)[2]
% ##  A: Yes (around 3/4 an act of protest difference)
% @

Messages left for your future self (or near-future others) help
retrace and justify your decisions. If you ever need to revist
them as the work moves from seminar paper to conference paper to
poster back to paper to dissertation and onwards. Notice one another
benefit of coding for an audience: we learn by teaching. By assuming
that others will look at your code, you will be more likely to write
clearer code, or perhaps even to learn more about what you are doing
as you do it.
